---
title: "R Assignment Two"
author: "Katherine Wolf"
date: "February 11, 2020"
output:
    pdf_document:
       includes:
         in_header: preamble-latex.tex
    html_document:
       includes:
         before_body: preamble-mathjax.tex
    latex_engine: xelatex
mainfont: Calibri
monofont: Lucida Console
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE)

```


```{r load required packages and read data}

library(foreign)
library(boot)
load("frmgham_recoded.wide.Rdata")

```

```{r}

# Make 2nd category (BMI 18.5-24.9, ideal weight) the referent group:
frmgham_recoded.wide$bmi_cat <- 
  relevel(as.factor(frmgham_recoded.wide$bmi_cat),"2")

```

```{r}

```



```{r}

standardized.measures <- function(dataset, index){
  # Create resampled version of dataset using index vector:
  dataset.resampled <- dataset[index,];

  # Fit logistic regression from the resampled dataset:
  # Outcome: hyperten
  # Covariates: bmi_cat, cursmoke, age, sex, educ
  # Store results in object called logistic.frmgham
  
  logistic.frmgham <- ##### COMPLETE ON OWN 
    glm(hyperten ~ bmi_cat + cursmoke + age + sex + educ,
        data = dataset.resampled, 
        family = binomial(link = "logit"))
  
  ##### end complete on own
  
  ### To calculate the measures of association:

  # STEP 1: Create two new versions of the *original* dataset, called
  # frmgham_recoded.wide.obese and frmgham_recoded.wide.ideal
  frmgham_recoded.wide.obese <- 
    frmgham_recoded.wide.ideal <- 
    frmgham_recoded.wide

  # Set BMI to obese (category 4) in population 1 [exposed]
  # and ideal weight (category 2) in population 0 [unexposed]:
  frmgham_recoded.wide.obese$bmi_cat <- "4" # population w/ all obese
  frmgham_recoded.wide.ideal$bmi_cat <- "2" # population w/ all ideal weight

  # STEP 2: Obtain predicted individual risk of hypertension under each new dataset:
  rhat.obese <- predict(logistic.frmgham, type = "response",
                        newdata = frmgham_recoded.wide.obese)
  rhat.ideal <- predict(logistic.frmgham, type = "response",
                        newdata=frmgham_recoded.wide.ideal)

  # STEP 3: Calculate the average risk (proportion) of hypertension
  # in each hypothetical population:
  risk.obese <- mean(rhat.obese)
  risk.ideal <- mean(rhat.ideal)
  
  # Calculate risk difference and risk ratio using ideal weight as reference:
  rd <- risk.obese - risk.ideal
  rr <- risk.obese/risk.ideal

  # STEP 4: Return these estimates:
  return(c(rd, rr))
}

```




```{r}

n.frmgham <- nrow(frmgham_recoded.wide)
stdized.measures <- standardized.measures(frmgham_recoded.wide, 
                                          index=seq(1:n.frmgham))
stdized.rd <- stdized.measures[1] # Risk difference is 1st element
stdized.rr <- stdized.measures[2] # Risk ratio is 2nd element

```



```{r}

set.seed(123)
# Put the bootstrapped sample results into object called bs.standardized
bs.standardized <- boot(frmgham_recoded.wide, standardized.measures, 5000)
boot.ci(bs.standardized, type= "bca", index=1)
boot.ci(bs.standardized, type= "bca", index=2)


```


```{r}

rd.samples <- bs.standardized$t[,1]  # risk difference
rr.samples <- bs.standardized$t[,2]  # risk ratio

par(mfrow=c(2,1))
plot(density(rd.samples), main = "Bootstrapped Samples of Risk Difference")
plot(density(rr.samples), main = "Bootstrapped Samples of Risk Ratio")
par(mfrow=c(1,1))

```


```{r on my own pretending}

# Calculate the standard deviation of each of the series of bootstrapped samples obtained in the above code chunk (use the log-scale for the relative measure). Calculate these centered on the original point estimate (not the sample means).

sd.rd.boot <- sd(rd.samples - stdized.rd)

sd.log.rr.boot <- sd(log(rr.samples) - log(stdized.rr))


# Use these estimates of standard deviation to calculate 95% confidence intervals for the RD and RR using a Normal approximation approach (see slide 12 from lecture notes). (Hint: make sure to do the CI calculation for the RR on the log scale and transform back to ratio scale.)

rd.ci.norm.apx <- c(stdized.rd - 1.96*sd.rd.boot, 
                    stdized.rd + 1.96*sd.rd.boot)

log.rr.ci.norm.apx <- c(log(stdized.rr) - 1.96*sd.log.rr.boot, 
                        log(stdized.rr) + 1.96*sd.log.rr.boot)

rr.ci.norm.apx <- exp(log.rr.ci.norm.apx)

```

# Question One

## State the point estimate and 95% confidence interval for the estimated risk difference and risk ratio for this analysis (20 points)

The estimated risk difference is `r stdized.rd` with a 95% confidence interval of ``

# Question Two

## Turn in the density plots for the risk difference and risk ratio measures. Briefly describe the pattern that you see. Does the pattern seem to indicate that the sampling distribution of the RD and RR are approximately normal? (10 points)

# Question Three

## Showing your work, calculate the 95% CI for the risk difference using a Normal approximation, and report your answer. (Show the formula you used, and the specific values that went into your calculation.) What assumptions does this require? How does this compare to the results from question one? (10 points)

# Question Four

## Showing your work, calculate the 95% CI for the risk ratio, and report your answer. (Show the formula you used, and the specific values that went into your calculation.) What assumptions does this require? How does this compare to the results from question one? (10 points)

# Appendix: R code

```{r ref.label=knitr::all_labels(), echo = T, eval = F}

```

